As a prelude, we start with an application of classical Gr\"obner bases that deals with a seemingly infinite problem; here, of course, a recurrence helps to ``control infinity''.  

The Fibonacci sequence $F_n = 1,1,2,3,5,\ldots$ ($n= 1, 2, \ldots$) is a \textit{strong divisibility sequence}, in that we have $\gcd(F_n, F_m) = F_{\gcd(n,m)}$; in particular, $F_m$ divides $F_n$ if $m \, | \, n$.  This surprising fact was used by \'Edouard Lucas for Mersenne prime testing.  Is there a direct explanation for the integrality of $F_{3n}/F_n \in \mathbb Z$?  It turns out that there is an identity:
\begin{equation}\label{F3n}
(F_{3n} - 5 F_n^3 - 3 F_n)(F_{3n} - 5 F_n^3 + 3 F_n) = 0,
\end{equation}
which explains in an explicit manner strong divisibility for this case.   Is it possible to use Gr\"obner bases to derive this identity?  The following Macaulay 2 code does exactly that:
\begin{M2}
\begin{verbatim}
i1: R = QQ[z, x, y, t, MonomialOrder => Eliminate 2]

i2: I = ideal(x + y - z, (x*z - y^2)^2 - 1, t - z^3 - y^3 + x^3)

i3: toString groebnerBasis I

o3 = matrix {{25*y^6-10*y^3*t-9*y^2+t^2, z-x-y, ...
\end{verbatim}
\end{M2}  
In this computation, the variables $z,x,y,t$ represent the recurrence values \mbox{$F_{n+1}, F_{n-1}, F_n, F_{3n}$,} respectively.  The first generator of $I$ defines the recurrence, the second is Cassini's identity, and the third is Lucas'.

One can check that factoring the first polynomial in the list above gives (\ref{F3n}).  Bootstrapping with extra equations, we can also discover that:
\begin{equation}\label{F5n}
(F_{5n} - 25 F_n^5 - 25 F_n^3 - 5 F_n)(F_{5n} - 25 F_n^5 + 25 F_n^3 - 5 F_n) = 0.
\end{equation}
In turn, these findings incite conjectures and proofs.  
For instance, we leave it to the reader to use modular arithmetic to verify from (\ref{F5n}) that the integer $\frac{F_{5n}}{5F_n}$ always has unit digit $1$ base ten.  
More generally, the following natural problem arises from this investigation:  Given $\ell$, find a nonzero polynomial $P(y, t) \in \mathbb Z[y, t]$ satisfying an identity of the form $P(F_n, F_{\ell n}) = 0$ (see \cite{HilLev:07} for more on ``polynomial recurrences").

The above is classical.  Here, we are interested in problems with not four or even twenty-four indeterminates, but rather an infinite number of them.  Take, for instance, the following basic ideal membership question.  Let $I \subset \mathbb C[x_0,x_1,\ldots]$ be the ideal generated by all permutations acting on the polynomial $f = x_0 x_1 - x_1 x_2^2 +x_1^2$.  Is the following in $I$?
\[ h = x_0 x_4^2 + x_0 x_1^2  +x_1 x_0^2 - 2 x_1 x_0 + x_0 x_3 x_4 - x_0 x_5^2 - x_0 x_3 x_5 - 2 x_1^2.\]

The difference in this question from classical problems of polynomial algebra is that \textit{a priori} there is no guarantee a particular computation, say, with a truncated polynomial ring $\mathbb C[x_0,x_1,\ldots, x_N]$ will do the job.  Nonetheless, the following code gives us an answer to our question \cite{EquivariantGB}.
\begin{M2}
\begin{verbatim}
i1: needsPackage "EquivariantGB"

i2: R = buildERing({symbol x}, {1}, QQ, 6);

i3: h = x_0*x_4^2+x_0*x_1^2+x_1*x_0^2-2*x_1*x_0+x_0*x_3*x_4- ...

i4: G = egb({x_0*x_1 - x_1*x_2^2 + x_1^2}, Algorithm=>Incremental)

       2       2      2           3      2     2    2         
o4 = {x x  - 2x  + x x  - 2x x , x  - x x , x x  - x  - x x , 
       1 0     1    1 0     1 0   1    1 0   2 0    1    1 0 

                    2           2 
      x x  - x x , x  + x x  - x  - x x }
       2 1    2 0   2    2 0    1    1 0

i5: reduce(h, G)

o5 = 0
\end{verbatim}
\end{M2}  
With the \EGB\ produced above, we can solve ideal membership problems and much more, just as we can use classical Gr\"obner bases in numerous applications.  %Note that arbitrarily large numbers of generators \cite{hillar2008minimal}.  

Developing the machinery to solve such questions is more than an intellectual curiosity.  Not only can basic facts now be proved by computer such as Theorem~\ref{toric2x2} but also cutting edge conjectures.  For example, using \cite{EquivariantGB}, it is possible to verify \cite{draisma2013noetherianity, Krone:egb-toric} the first nontrivial case of a basic finiteness conjecture for toric ideals \cite{aschenbrenner2007finite}.  

\begin{theorem}[Proved by computer]\label{monomthm}
For $n > 1$, let $I_n = \ker (y_{ij} \mapsto x_i^2 x_j)$, $1 \leq i \neq j \leq n$.  The invariant chain of toric ideals $I_2 \subset I_3 \subset \cdots$ stabilizes up to the symmetric group.  That is, there is some $N$ such that all elements of $I_m$, $m > N$, are polynomial consequences of relabellings of a finite set of generators of $I_N$.
\end{theorem}

%We note that whether there is always an equivariant Gr\"obner bases is still an open question.

We next provide a summary of applications of \EGBs\ in rings with infinite numbers of indeterminates.  

\subsubsection{Group theory and Chemistry.}

The first use of the concept ``finite up to symmetry" for polynomial rings that we are aware of is in the group theory work of Cohen in \cite{cohen1967laws}.  
Independently, it was problems in algebraic chemistry \cite{ruch1967vandermondesche}, brought to the attention of the authors of \cite{aschenbrenner2007finite} by Andreas Dress, that motivated further applications of asymptotic polynomial algebra in chemistry \cite{Draisma08b}.

\subsubsection{Toric Algebra and Algebraic Statistics.}

A major inspiration for asymptotic algebra arises from studying chains of toric ideals, many of which arise naturally in algebraic statistics.  The series of works \cite{Hillar13, hillar2016corrigendum, draisma2013noetherianity, KKL:equivariant-markov, Krone:egb-toric} have developed fundamental finiteness properties of these structures, but many questions remain open, as we outline in Section \ref{sec:challenges}.  

In this regard, one of the major motivations for \EGBs\ and infinite symbolic algebra are their application to the problem of sampling from conditional distributions by algebraic methods \cite{diaconis1998algebraic}.  At its essence, the strategy is to find a collection of elementary moves through model space that preserves the sufficient statistics of the data.   The idea then is to consider growing families of model classes and show that, up to obvious symmetries, only a finite set of moves suffices for all infinite numbers of models (e.g., \cite{aoki2003minimal, santos2003higher, hocsten2007finiteness, drton2007algebraic, Draisma08b, Brouwer09e, draisma2009ideals, hillar2012finite, draisma2015finiteness}).  Typically, these moves correspond to elements of a Gr\"obner basis or at least a generating set for some ideal.

\subsubsection{Invariants.}
Recently, Nagel and R\"omer~\cite{Nagel} introduced Hilbert series for Noetherian infinite-dimensional rings. Their original theoretical treatment that leads to a proof of rationality of the Hilbert series, in principle, also leads to an effective procedure to compute the series. For an ideal generated up to symmetry by one monomial, such computation was carried out in~\cite{gunturkun2016equivariant}. An alternative approach of~\cite{krone2016hilbert} computes the Hilbert series given an equivariant Gr\"obner basis as the generating function counting words in a regular language.  

\subsection{Finiteness up to symmetry in general.}
Although the \EGBs\ described in this article may not directly apply, finiteness up to symmetry plays the central role in the following results (this list is by no means exhaustive).  

It appears in homological stability~\cite{randal2013homological, church2012homological},  the moduli space of $n$ points in a line~\cite{howard2009equations}, geometry as the positivity of the the embedding line bundle grows \cite{ein2012asymptotic}, syzygies of Segre embeddings~\cite{snowden2013syzygies}, Betti tables as their length goes to infinity~\cite{ein2015asymptotics}, tensor geometry~\cite{draisma2014bounded, draisma2015finiteness}, and limiting Grassmannians~\cite{draisma2015plucker}.
Gr\"obner methods have also been used to understand representations of combinatorial categories \cite{sam2016grobner}.
