\documentclass[11pt]{amsart}
\usepackage{amssymb}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{color}
\usepackage{soul}
\usepackage{setspace}
\usepackage{graphicx}

\newenvironment{M2}{ \begin{spacing}{0.4} %\begin{quote}
\bigskip \small} { %\end{quote}
\end{spacing} 
\bigskip 
}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}\newtheorem{conjecture}[theorem]{Conjecture}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{xca}[theorem]{Exercise}
\newtheorem{algorithm}[theorem]{Algorithm}

%\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{criterion}[theorem]{Criterion}

\numberwithin{equation}{section}

\newcommand{\B}[1]{\mathbb #1}
\newcommand{\C}[1]{\mathcal #1}
\newcommand{\F}[1]{\mathfrak #1}
\newcommand{\RM}[1]{\mathrm #1}
\newcommand{\SF}[1]{\mathsf #1}
\newcommand{\BF}[1]{\mathbf #1}

\newcommand{\alg}[1]{\operatorname{#1}}
\newcommand{\ideal}[1]{\langle #1 \rangle}
\newcommand{\norm}[1]{\| #1 \|_1}
%\newcommand{\dfrac}{\genfrac{}{}{}0}
%\newcommand{\tfrac}{\genfrac{}{}{}1}

\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SU}{SU}
\DeclareMathOperator{\Id}{Id}
\DeclareMathOperator{\poly}{poly}
\DeclareMathOperator{\initial}{in}
\DeclareMathOperator{\NF}{NF}
\DeclareMathOperator{\lcm}{lcm}
\DeclareMathOperator{\im}{im}
\DeclareMathOperator{\Syz}{Syz}
\newcommand{\dist}{d}
\newcommand{\actson}{\curvearrowright}
\newcommand{\cover}{\widetilde}
\newcommand{\closure}{\overline}
\newcommand{\Inc}{\operatorname{Inc}(\B N)}
\newcommand{\IncT}{\operatorname{Inc}^\tau(\B N)}
\newcommand{\SymN}{\F S_\infty}
\newcommand{\Sn}{\tilde{\F S}_n}
\newcommand{\Snp}{\tilde{\F S}_{n'}}
\newcommand{\Sinf}{\F S_{\infty}^r}
\newcommand{\lleq}{\preceq}
\newcommand{\ggeq}{\succeq}
\newcommand{\llleq}{\sqsubseteq}
\newcommand{\gggeq}{\sqsupseteq}
\newcommand{\ZS}{\B Z^{\C S}}
\newcommand{\ZT}{\B Z^{\C T}}
\newcommand{\ZSn}{\B Z^{\C S_n}}
\newcommand{\ZTn}{\B Z^{\C T_n}}
\newcommand{\ZSone}{\B Z^{\C S_{(1,\ldots,1)}}}
\newcommand{\ZSC}{\ZS_{\mathbf{sc}}}
\newcommand{\Lin}{\mathrm{Lin}}
\newcommand{\mon}{M}
\newcommand{\Sym}{\F S_\infty}
\newcommand{\LT}{\initial_{\leq}}
\newcommand{\LTp}{\initial_{\leq'}}
\newcommand{\LC}{\operatorname{LC}}
\newcommand{\LM}{\operatorname{LM}}
\newcommand{\mm}{\C M}
\newcommand{\degtot}{\operatorname{deg}_{tot}}
\newcommand{\NN}{\BF N}


\newcommand{\GB}{{Gr\"obner basis}}
\newcommand{\GBs}{{Gr\"obner bases}}
\newcommand{\EGB}{{equivariant Gr\"obner basis}}
\newcommand{\EGBs}{{equivariant Gr\"obner bases}}

\begin{document} \title{Equivariant Gr\"obner bases}

 %   Information for first author 
\author{Christopher J. Hillar}
\address{University of California, Berkeley}
\email{chillar@msri.org}

\author{Robert Krone}
\address{Queens University}
\email{rk71@queensu.ca}

\author{Anton Leykin}
\address{Georgia Tech}
\email{leykin@math.gatech.edu}
\thanks{Research of AL is supported in part by NSF grant DMS-1151297.}

\thanks{Part of this work took place during the ``Free Resolutions, Representations, and Asymptotic Algebra'' workshop at the BANFF International Research Station, April, 2016.} 

%\subjclass{13E05, 13E15, 20B30, 06A07}

%\keywords{Invariant ideal, well-quasi-ordering, symmetric group, Gr\"obner basis, generating sets, infinite dimensional algebra}

% ----------------------------------------------------------------
\begin{abstract}
Algorithmic computation in polynomial rings is a classical topic in mathematics. However, little attention has been given to the case of rings with an infinite number of variables until recently when theoretical efforts have made possible the development of effective routines. Ability to compute relies on finite generation up to symmetry for ideals invariant under a large group or monoid action, such as the group of permutations of the natural numbers. We summarize the current state of theory and applications for equivariant Gr\"obner bases, develop several algorithms to compute them, and showcase our software implementation. 
\end{abstract} 

\maketitle 
%\tableofcontents
\section{Introduction}
\subsection{History}

\input{history.tex}

\subsection{Applications}

\input{applications.tex}

\subsection{Goals and structure}

Since their introduction, Gr\"obner bases techniques have improved immensely.  We believe that in this new setting of infinite-dimensional polynomial algebras, which demands far more computational power, algorithmic development is at the start of a similar road. We hope a similar progress in theoretical understanding and in computing techniques occurs in the future.  Our goals are to outline the current state of effective computation in this setting and provide a background for  researchers to start to tackle this exciting new computational domain.  

After some preliminaries in Section~\ref{prelim}, we quickly move to describing equivariant Gr\"obner bases algorithms in Section~\ref{EGB}.  Section~\ref{sec:signature} goes on to explain a modern signature-based approach and a strategy inspired by it for an equivariant Buchberger's algorithm. The final Section~\ref{sec:challenges} outlines computational and theoretical challenges for future exploration.

\section{Preliminaries}\label{prelim}

\input{preliminaries.tex}

\section{Equivariant Buchberger algorithm}\label{EGB}

\input{e-buchberger.tex}

\section{A signature-based approach}\label{sec:signature}

\input{egb-signature.tex}

\section{Open questions and challenges}\label{sec:challenges}

\input{challenges.tex}


\bibliographystyle{plain}
\bibliography{egb}

\end{document}
% ----------------------------------------------------------------
`
